---
title: "Tutorial Part IV - Movement - Exercise 2"
author: "Stephanie Kramer-Schadt & Cédric Scherer"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y'))`"
output:
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}
p {
  line-height:170%;
}
#sidebar h2 {
  background-color: Indigo;
}
code {
  color: Indigo ;
}
.exercise {
  color: #824b00;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 9, fig.height = 6, dev = "ragg_png")
#knitr::opts_knit$set(root.dir = 'C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer')
#or on GUI: knit -> knit directory -> project directory
```

#  Course 4

## Exercise 2

**Animals have been collared and tracked, and we want to know their home range size and whether the animals actually avoid agriculture and prefer forest. <br>
There is telemetry data available from three animals (`animal_relocations_32633.shp`) in the
folder data_move. Additionally, land cover (`landuse_3035.shp`) and rivers (`water_linear_32633.shp`) are available in the `data_move/geo-raw` folder. 
Land use has a column `NS1` with coded info on `1` (urban areas), `2` (agriculture), 
`3` (forest), `4` (swamps), and `5` (water bodies).<br>**

After loading the necessary libraries, the data and making a plot, please do the following:<br>

* a. Calculate their home ranges. Hint: set the argument for multiple animals in move()<br>
* b. Buffer the home range with the mean steplength of all three animals.<br>
* c. Clip the home ranges with the underlying landscape and calculate the percentage of land use classes 2 (agriculture) and 2 (forest) within the home range (hint: use `st_intersection()`  AND
recalculate the column for AREA in km2).<br>
* d. In which land use are the animals actually located? Extract land use underneath each telemetry location (hint: `st_join()`). Do the animals actually avoid agriculture?
<br><br>

<br><hr><br>

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
sessionInfo()
```

</details>
