---
title: "Tutorial Part VI — homerange estimates with amt"
author: "Stephanie Kramer-Schadt & Moritz Wenzler-Meya"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y'))`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: show
    toc_depth: 4
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1, h2, h3 {
  margin-top: 50px;
}
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}
p {
  line-height:170%;
}
sidebar h2 {
  background-color: Indigo;
}
code {
  color: Indigo ;
}
.exercise {
  color: #824b00;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      dev = "ragg_png", fig.width = 9, fig.height = 6, dpi = 600, retina = 1)
Sys.setlocale("LC_TIME", "C")
```

<span class='exercise'> Course 4 homerange estimates with amt </span>


# Preparations

## Load libraries

```{r load packages,message=FALSE}

### The packages (repetition) ###
   library(here)
   library(sf)
   library(tmap)
   library(terra)
   library(amt)
   library(viridis)
   #library(ctmm)
   #library(adehabitatHR)
```

## Set the workspace

```{r workspace}
### The workspace  (repetition) ###
getwd() # you can also use the package 'here()'

# my data are outside my course-folder, therefore I have to do it the old-fashioned way.
root_wd <- here::here() 
# relative to work-wd
maps_wd <- paste(root_wd,"/","data/data_move",sep='') # or:
geo_wd <- paste(root_wd,"/","data/data_move/geo-raw",sep='')

# the output folder should have been created by you during Tutorial 2 'R goes spatial'.
# It should contain the hillshade.asc
output_wd <- here::here("output")

setwd(output_wd) ## set to the OUTPUT folder!
getwd() # check
```

# Load spatial data and define the CRS

## animal locations

```{r animal locations}
### Data import - Load spatial data (repetition) ###

animal_relocations <- sf::st_read(paste(maps_wd, "animal_relocations_32633.shp", sep = "/"))

head(animal_relocations)
```

## rivers

```{r rivers}
rivers <- sf::st_read(paste(geo_wd, "water_linear_32633.shp", sep = "/"))

head(rivers)
```


```{r, eval=FALSE}
sf::st_crs(rivers) # check the crs
```

```{r}
sf::st_crs(rivers)$epsg # show the epsg code
```


## landuse 

```{r landuse}
landuse <- sf::st_read(paste(geo_wd, "landuse_3035.shp", sep = "/"))

head(landuse)

#sf::st_crs(landuse) # check crs
sf::st_crs(rivers)$epsg # show the epsg code

# transform to river crs
landuse_trans <- sf::st_transform(landuse, sf::st_crs(rivers))
```

## kettle holes

```{r kettle holes}
kettleholes <- read.delim(paste(maps_wd, "KettleHoles2.txt", sep = "/"))

head(kettleholes)

# tranform to sf object
kettleholes_sf <- sf::st_as_sf(x = kettleholes,
                               coords = c("MyX", "MyY"),
                               crs = 32633) # which crs should we use? Compare coordinates with coordinates from the other files
```

## plot the data with tmap

```{r,out.height="700px",out.width="700px"}
tmap::tmap_mode("view")
tmap::tmap_options(check.and.fix = TRUE)
tmap::tm_shape(landuse_trans) + tmap::tm_polygons(col = "NS1", palette = "Greens") +
tmap::tm_shape(animal_relocations) + tmap::tm_dots(col = "red") + 
  tmap::tm_shape(rivers) + tmap::tm_lines(col = "blue") + 
  tmap::tm_shape(kettleholes_sf) + tmap::tm_dots(col = "orange")

```

# Biodiversity index

## create buffers around kettleholes

```{r}
kettleholes_sf_buf <- sf::st_buffer(x = kettleholes_sf, dist = 1000)# test different buffers (e.g. 100m, 500m, 1000m....)

head(kettleholes_sf_buf)
```

## intersect landuse with kettleholes

```{r}
landuse_ext_buf <- sf::st_intersection(landuse_trans, kettleholes_sf_buf)

head(landuse_ext_buf)
```

## calculate area of each category of column NS1

```{r}
landuse_ext_buf$area_ns1 <- sf::st_area(landuse_ext_buf)

# show the first 20 area calculations of the landuse category
head(landuse_ext_buf$area_ns1, 20)

# check first which categories are existing in the data

table(landuse_ext_buf$NS1)

landuse_df <- data.frame(landuse_ext_buf) # transform to a data frame
landuse_df <- landuse_df[,c("No","NS1","area_ns1")] # select only necessary columns

landuse_df_sub <- data.frame(
  No = c(1,2,3), # add No column by Hand
  NS1 = c(1,2,3), # add NS1 column category by Hand
  # calculate the area per No and NS1 category
  area_ns1_1 = c(sum(landuse_df[which(landuse_df$No == 1 & landuse_df$NS1 == 1 ),]$area_ns1), # No = 1 and NS1 = 1
               sum(landuse_df[which(landuse_df$No == 2 & landuse_df$NS1 == 1 ),]$area_ns1), # No = 2 and NS1 = 1
               sum(landuse_df[which(landuse_df$No == 3 & landuse_df$NS1 == 1 ),]$area_ns1) # No = 3 and NS1 = 1
               ),
  area_ns1_2 = c(sum(landuse_df[which(landuse_df$No == 1 & landuse_df$NS1 == 2 ),]$area_ns1), # No = 1 and NS1 = 2
               sum(landuse_df[which(landuse_df$No == 2 & landuse_df$NS1 == 2 ),]$area_ns1), # No = 2 and NS1 = 2
               sum(landuse_df[which(landuse_df$No == 3 & landuse_df$NS1 == 2 ),]$area_ns1) # No = 3 and NS1 = 2
               ),
  area_ns1_3 = c(sum(landuse_df[which(landuse_df$No == 1 & landuse_df$NS1 == 3 ),]$area_ns1), # No = 1 and NS1 = 3
               sum(landuse_df[which(landuse_df$No == 2 & landuse_df$NS1 == 3 ),]$area_ns1), # No = 2 and NS1 = 3
               sum(landuse_df[which(landuse_df$No == 3 & landuse_df$NS1 == 3 ),]$area_ns1) # No = 3 and NS1 = 3
               )
)


# combine kettleholes sf data set with the extracted landuse data
kettleholes_sub <- dplyr::left_join(kettleholes_sf, landuse_df_sub, by = "No")

head(kettleholes_sub)
```

## intersect the animal relocations with the landuse

```{r}
animal_relocations_inter <- sf::st_intersection(animal_relocations, landuse_trans[,"NS1"])

head(animal_relocations_inter)

# compare landuse category per animal ID
table(animal_relocations_inter$NS1, animal_relocations_inter$ID)
```


# mcp and kde with amt

## data preperation for amt package

```{r}
## split by animal id

animal_relocations_id1 <- animal_relocations[which(animal_relocations$ID == 1),] 
animal_relocations_id2 <- animal_relocations[which(animal_relocations$ID == 2),]
animal_relocations_id3 <- animal_relocations[which(animal_relocations$ID == 3),]

# transform the sf data to a data frame for later use with the amt package.

animal_relocations_df_id1 <- data.frame(id = animal_relocations_id1$ID,
                                     x = sf::st_coordinates(animal_relocations_id1)[,1], # get the coordinates. First column for x
                                     y = sf::st_coordinates(animal_relocations_id1)[,2], # get the coordinates. Second column for y
                                     t = 1:nrow(animal_relocations_id1)) # add t column with the row id 

animal_relocations_df_id2 <- data.frame(id = animal_relocations_id2$ID,
                                     x = sf::st_coordinates(animal_relocations_id2)[,1], # get the coordinates. First column for x
                                     y = sf::st_coordinates(animal_relocations_id2)[,2], # get the coordinates. Second column for y
                                     t = 1:nrow(animal_relocations_id2))

animal_relocations_df_id3 <- data.frame(id = animal_relocations_id3$ID,
                                     x = sf::st_coordinates(animal_relocations_id3)[,1], # get the coordinates. First column for x
                                     y = sf::st_coordinates(animal_relocations_id3)[,2], # get the coordinates. Second column for y
                                     t = 1:nrow(animal_relocations_id3))
```

## calculate mcp kettlehole 1

Create an MCP and an KDE for each kettlehole seperatly. Compare the MCP and KDE. What is the difference?

```{r}
# convert to track object. This is necessary to work with the amt package
animal_relocations_hr_id1 <- amt::make_track(tbl = animal_relocations_df_id1,
                                          .x = x,
                                          .y = y,
                                          .t = t,
                                          crs = sf::st_crs(animal_relocations)) # we are using the same crs as in the original data

mcps_id1 <- amt::hr_mcp(x = animal_relocations_hr_id1, # the track object
                        levels = c(0.95)) # define the level of the mcp. How many data points are covered by the mcp

mcps_sf_id1 <- sf::st_as_sf(mcps_id1$mcp) # convert to sf object to use it outside of the amt package

mcps_sf_id1
```

## calculate kde for kettlehole 1

```{r}
kde_id1 <- amt::hr_kde(x = animal_relocations_hr_id1, # the track object
                       levels = c(0.95)) # as in mcp

kde_sf_id1 <- amt::hr_isopleths(kde_id1) # get the isopleths of the kde to have a polygon

kde_sf_id1
```

## create random points within the mcp

```{r}
random_pts_id1 <- sf::st_sample(x = mcps_sf_id1, # as input we use the mcp polygon
                                size = nrow(animal_relocations_id1)) # same number as locations

random_pts_sf_id1 <- sf::st_as_sf(random_pts_id1) # transform to a proper sf object

head(random_pts_sf_id1)
```

## intersect the random points with the landuse dataset

```{r}
random_pts_id1_inter <- sf::st_intersection(random_pts_sf_id1, # input random points
                                            landuse_trans[,"NS1"]) # transformed landuse 

random_pts_id1_inter$ID <- 1 # add ID

head(random_pts_id1_inter)
```

## calculate mcp for kettlehole 2

Create an MCP and an KDE for each kettlehole seperatly. Compare the MCP and KDE. What is the difference?

```{r}
# convert to track object. This is necessary to work with the amt package
animal_relocations_hr_id2 <- amt::make_track(tbl = animal_relocations_df_id2,
                                          .x = x,
                                          .y = y,
                                          .t = t,
                                          crs = sf::st_crs(animal_relocations)) # we are using the same crs as in the original data

mcps_id2 <- amt::hr_mcp(x = animal_relocations_hr_id2, # the track object
                        levels = c(0.95)) # define the level of the mcp. How many data points are covered by the mcp

mcps_sf_id2 <- sf::st_as_sf(mcps_id2$mcp) # convert to sf object to use it outside of the amt package
```

## calculate kde for kettlehole 2

```{r}
kde_id2 <- amt::hr_kde(x = animal_relocations_hr_id2, # the track object
                       levels = c(0.95)) # as in mcp

kde_sf_id2 <- amt::hr_isopleths(kde_id2) # get the isopleths of the kde to have a polygon

kde_sf_id2
```

## create random points within the mcp

```{r}
random_pts_id2 <- sf::st_sample(x = mcps_sf_id2, # as input we use the mcp polygon
                                size = nrow(animal_relocations_id2)) # same number as locations

random_pts_sf_id2 <- sf::st_as_sf(random_pts_id2) # transform to a proper sf object

head(random_pts_sf_id2)
```

## intersect the random points with the landuse dataset

```{r}
random_pts_id2_inter <- sf::st_intersection(random_pts_sf_id2, # input random points
                                            landuse_trans[,"NS1"]) # transformed landuse 

random_pts_id2_inter$ID <- 2 # add ID

head(random_pts_id2_inter)
```

## calculate mcp for kettlehole 3

Create an MCP and an KDE for each kettlehole seperatly. Compare the MCP and KDE. What is the difference?

```{r}
# convert to track object. This is necessary to work with the amt package
animal_relocations_hr_id3 <- amt::make_track(tbl = animal_relocations_df_id3,
                                          .x = x,
                                          .y = y,
                                          .t = t,
                                          crs = sf::st_crs(animal_relocations)) # we are using the same crs as in the original data

mcps_id3 <- amt::hr_mcp(x = animal_relocations_hr_id3, # the track object
                        levels = c(0.95)) # define the level of the mcp. How many data points are covered by the mcp

mcps_sf_id3 <- sf::st_as_sf(mcps_id3$mcp) # convert to sf object to use it outside of the amt package

mcps_sf_id3
```

## calculate kde for kettlehole 3

```{r}
kde_id3 <- amt::hr_kde(x = animal_relocations_hr_id3, # the track object
                       levels = c(0.95)) # as in mcp

kde_sf_id3 <- amt::hr_isopleths(kde_id3) # get the isopleths of the kde to have a polygon

head(kde_sf_id3)
```

## create random points within the mcp

```{r}
random_pts_id3 <- sf::st_sample(x = mcps_sf_id3, # as input we use the mcp polygon
                                size = nrow(animal_relocations_id3)) # same number as locations

random_pts_sf_id3 <- sf::st_as_sf(random_pts_id3) # transform to a proper sf object

head(random_pts_sf_id3)
```

## intersect the random points with the landuse dataset

```{r}
random_pts_id3_inter <- sf::st_intersection(random_pts_sf_id3, # input random points
                                            landuse_trans[,"NS1"]) # transformed landuse 

random_pts_id3_inter$ID <- 3 # add ID

head(random_pts_id3_inter)
```

## plot the data

```{r,out.height="700px",out.width="700px"}
tmap::tmap_mode("view")
tmap::tm_shape(mcps_sf_id1) + tmap::tm_borders(col = "blue") + 
  tmap::tm_shape(kde_sf_id1) + tmap::tm_borders(col = "orange") + 
  tmap::tm_shape(animal_relocations_id1) + tmap::tm_dots(col = "green") + 
  tmap::tm_shape(random_pts_id1_inter) + tmap::tm_dots(col = "red") +
tmap::tm_shape(mcps_sf_id2) + tmap::tm_borders(col = "blue") + 
  tmap::tm_shape(kde_sf_id2) + tmap::tm_borders(col = "orange") + 
  tmap::tm_shape(animal_relocations_id2) + tmap::tm_dots(col = "green") + 
  tmap::tm_shape(random_pts_id2_inter) + tmap::tm_dots(col = "red") +
tmap::tm_shape(mcps_sf_id3) + tmap::tm_borders(col = "blue") + 
  tmap::tm_shape(kde_sf_id3) + tmap::tm_borders(col = "orange") + 
  tmap::tm_shape(animal_relocations_id3) + tmap::tm_dots(col = "green") + 
  tmap::tm_shape(random_pts_id3_inter) + tmap::tm_dots(col = "red") 
```

## combine all random points

```{r}
random_pts_inter_all <- rbind(random_pts_id1_inter,
                              random_pts_id2_inter,
                              random_pts_id3_inter)
```


# calcualte xhi square test

```{r}
table(random_pts_inter_all$NS1) # random
table(animal_relocations_inter$NS1) # observed

#LANDNUTZUNGSTYP
# TYP             #Feld(NS1=2)      #Wald(NS1=3)
#Angebot(random)     278              110
#Tier (observed)     168              220

chi_matrix <- matrix(data=c(278,110,168,220),ncol=2)
chisq.test(chi_matrix)

# 'von Hand' ausrechnen:
aa <- chi_matrix[1,1]
bb <- chi_matrix[1,2]
cc <- chi_matrix[2,1]
dd <- chi_matrix[2,2]

Na <- aa+bb
Nb <- cc+dd
Ns <- aa+cc
Nf <- bb+dd
N <- Na+Nb

mychi <- (N*(abs((aa*dd) - (bb*cc)) - (N/2))^2) / (Ns*Nf*Na*Nb)
mychi
```


<br><hr><br>

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
#git2r::repository() ## uncomment if you are using GitHub
sessionInfo()
```

</details>
