---
title: "Tutorial Part III a — SDM with MaxEnt"
author: "Stephanie Kramer-Schadt, Juergen Niedballa, Moritz Wenzler"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y'))`"
output:
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}
p {
  line-height:170%;
}
sidebar h2 {
  background-color: Indigo;
}
code {
  color: Indigo ;
}
.exercise {
  color: #824b00;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 9, fig.height = 6, dev = "ragg_png")
#knitr::opts_knit$set(root.dir = 'C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer/_GitHub')
```

<span class='exercise'> Course 3 SDM with MaxEnt </span>


# Preparations

## Load libraries

```{r,message=FALSE}

### The packages (repetition) ###
   library(here)
   library(sf)
   library(tmap)
   library(dismo)
   library(terra)
   library(viridis)
   library(rgl)
   library(unmarked)
   library(spatstat)

## check that the package raster is detached if you work with terra!
## However, it is later needed for the stack
# detach(package:raster)

#install.packages('rJava') ## needed only if you run MAxEnt from R
library(rJava)


```

## Set the workspace

```{r}
### The workspace  (repetition) ###
getwd() # you can also use the package 'here()'

root_wd <- here::here() 
# relative to work-wd


# TODO - Moritz' path adjust
#maps_wd <- paste(root_wd,"/","data/data_borneo/geo_raster_current_asc/selection_cur",sep='') # or:
#maps_wd_fut <- paste(root_wd,"/","data/data_borneo/geo_raster_future_asc/selection_fut",sep='')

maps_wd <- paste(root_wd,"/","data/data_borneo/geo_raster_current_asc",sep='') # or:
maps_wd_fut <- paste(root_wd,"/","data/data_borneo/geo_raster_future_asc",sep='')
vecs_wd <- here::here("data","data_borneo","geo_vector") # shapefile
recs_wd <- here::here("data","data_borneo","animal_data") # location data

# the output folder should have been created by you during Tutorial 2 'R goes spatial'.
# It should contain the hillshade.asc
output_wd <- here::here("output")

setwd(output_wd) ## set to the OUTPUT folder!
getwd() # check
```

## Load spatial data and define the CRS

### Load rasters

```{r}
### Data import - Load spatial data (repetition) ###

ras1 <- terra::rast(x = paste(maps_wd,'/bio_asc_01.asc',sep=''))
# assign the projection (crs - coordinate reference system)
# ras1@crs <- CRS("+proj=longlat +datum=WGS84") ## not needed in terra!

ras24 <- terra::rast(here("data","data_borneo", "geo_raster_current_asc", "bio_asc_24.asc")) #DEM

ras42 <- terra::rast(x = here("data", "data_borneo", "geo_raster_current_asc", "bio_asc_42.asc")) # land use

#hillsh <- terra::rast(x = paste(maps_wd,'/borneo_hillshade.asc',sep='')) 
```

### Stack environmental variables

Loading many rasters can be done in one go.
```{r}
### Raster stack of environmental predictors (repetition) ###

# the list.files command is very helpful to check what is in the folders
# use 'pattern' for searching for special file types, here .asc-files:
files <- list.files(path= maps_wd, pattern='.asc$',
                    full.names=TRUE )
files # these are just names! To load them as spatial objects, use rast()
## note that I only load 4 rasters:
# predictors <- c(files[c(9,12,22,24)]) # for full data set
predictors <- terra::rast(x = c(files[c(1,5,6,7)]) ) # for github repository data

predictors
```

### Plot rasters
```{r}
terra::plot(predictors, col = viridis::viridis(100)) # might take some time depending on your computer

# what are the differences? Have a look also at the data description file in the data folder.
```

### Load shapefiles

```{r,message=FALSE}
### Read in some Shapefiles (repetition) ###
## package sf (new and maintained)

## Borneo outline polygon
Borneo_shp_sf <- sf::st_read(dsn = vecs_wd, 
                      layer = "borneo_admin",
                      stringsAsFactors = FALSE)[,c(1:3,5,7,17,18)]
# Protected areas (PA) polygon
PA_shp_sf <-  sf::st_read(dsn = vecs_wd, 
                   layer = "Bor_PA",
                   stringsAsFactors = FALSE)[, c(1:4)]
# Rivers lines
River_shp_sf <- sf::st_read(dsn = vecs_wd, 
                     layer = "sn_100000",
                     stringsAsFactors = FALSE)
```

### Load point df and convert to spatial object

These are our observations of the species.
```{r}
### The spatial point data (species records) (repetition) ###

# filename
spec_pt_filename <- list.files(path = recs_wd, 
                               pattern = ".csv",
                               full.names = TRUE)#paste(recs_wd,'/','viverra_tangalunga.csv', sep='')
spec_pt_filename

# read the file
sp_recs <- read.csv(file = spec_pt_filename[1], # decide which species you want to use, here 1st in list
                    header=TRUE, sep=',')

#convert it to spatial object (sf here)
sp_recs_sf <- sf::st_as_sf(x = sp_recs, 
                       coords = c("long","lat"), # columns  for the coordinates
                       crs = 4326, # define crs, 4326 is the EPSG code
                       sf_column_name = "geometry",
                       remove=F) # sf needs a geometry column and you have to name it

```

## Plot data overview

```{r}
### Plot to get an impression ###

plot(ras42, col=grey.colors(20))
plot(PA_shp_sf$geometry,border='green', lwd=1.8, add=T)
plot(sp_recs_sf$geometry, pch= '*',cex=1,col='deeppink',add=T)
text(112.3, 6, 'Starting with SDMs', cex=1.2, col= 'red')

```

# Preconditions 

## Checking for multicollinearity

```{r}
# workaround for slow computers. First, aggregate the 1 km² resolution into 50*50 km  
agg_pred <- terra::aggregate(x=predictors,fact=50,FUN=mean)
plot(agg_pred)
```

### Pairs plot 

```{r}
terra::pairs(agg_pred, method = 'spearman')
```


```{r}
# Plot the differences (residuals) between the rasters:
diff <- terra::focalPairs(x = agg_pred, w = 3, 'pearson', na.rm = TRUE)
plot(diff)
```


```{r}
# not_run on slow computers:
#pairs(predictors)
```

NB: I am a big fan of the pairs()-plot because you actually see the distribution of the data, but you can do really nice plots with the `{corrplot}`-package: <br>
https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html

## Sampling bias correction

We are now creating a very rudimentary bias correction file. Let's assume that doing field work in the province of Sabah, Malaysia, is easy and the field sampling has concentrated on these areas. Further, it is easier to sample in lowlands and flat slopes. Usually, protected areas (PAs) are also well studied areas, where the sampling intensity is high. And let's assume that it was not possible to get a research permit for Indonesia and that therefore very few field studies had been conducted here. Please note: this is a completely fictive example! You should have the information about the sampling bias.

That is, we create a file that is representing these differences in sampling effort. This is easiest to do with raster manipulation. We first create a raster from our shapefile with the country outlines. 

```{r}
head(Borneo_shp_sf)
plot(Borneo_shp_sf[,3]) # column 3 is NAME_0 = main country names
```

We create a raster that gets the value of 1 for Malaysia (and Brunei) and a value of 0.01 (MaxEnt cannot deal with 0 -> see lecture) for the rest. We use one of our rasters as mask for cell size resolution and cropping extent:

```{r}
### *Rasterize* the shapefiles
## command rasterize() 
## field = 1 means all raster cells get value = 1, other cell values are set to 0
## background sets all other cells to a chosen number


Mal_ras <- terra::rasterize(x=Borneo_shp_sf[Borneo_shp_sf$NAME_0 %in%
                       c("Brunei","Malaysia"),], y=ras24, field=1,
                       background=0.1)

plot(Mal_ras)

# almost same as not plotting Indonesia:
# NotInd_ras <- rasterize(x=Borneo_shp_sf[Borneo_shp_sf$NAME_0 !=
#                           "Indonesia",], y=ras24, field=1, background=0.01)
```

We do the same only for the province of Sabah and the Protected Areas

```{r}
Sab_ras <- terra::rasterize(x=Borneo_shp_sf[Borneo_shp_sf$NAME_1 == 'Sabah',],
                     y=ras24, field=1, background=0.1)

PA_ras <- terra::rasterize(x=PA_shp_sf, y=ras24, field=1, background=0.1)


### Crosscheck with plot
par(mfrow=c(1,2))
plot(Sab_ras)
plot(PA_ras) #n.b.: only 1 and small values, no NA
par(mfrow=c(1,1))
```

And we do the same for all elevations below 300 m:
```{r}
ras24_300 <- ras24 <= 300
plot(ras24_300)
```

Now we add them all up. This means, we rank the sampling intensity, i.e. sampling in protected areas in Sabah below 300 m gets the highest value

```{r}
bias_tmp <- ras24_300 + Sab_ras + Mal_ras + PA_ras
bias_tmp
plot(bias_tmp)
```

Let's standardize between 0 and 1 (and then set the zeros to 0.01 again)
```{r}
# get the maximum value in the layer, standardize it and round to two decimals
maxval <- max(values(bias_tmp),na.rm=T)
bias_tmp2 <- bias_tmp/maxval
bias_tmp3 <- round(bias_tmp2, digits=2)

## same as in one go:   
bias1 <- round(bias_tmp/max(values(bias_tmp),na.rm=T),digits=2)

table(values(bias1)) 
# table(bias1@data@values) # is doing the same

### in case of having 0 somewhere:
bias1[values(bias1 == 0)] <- 0.01 # because MaxEnt
                                  # does not take 0!
table(values(bias1))
```

Now compare the rasters - do they have the same extent, cell size etc?
```{r}
terra::compareGeom(ras24,bias1) # the same?
```

check for the length of the NODATA values:
```{r}
# Are they really the same???
length(which(!is.na(values(bias1))))
length(which(!is.na(values(ras24))))
```
Nice! Now plot it:

```{r}
### Plot the bias file
plot(bias1, col = viridis(7)) #or col = grey.colors(7)
```

### Save the bias file

```{r}
## save into Output folder, which is your working directory: 
## check with getwd() if you're not sure of wd

## in case you want it aggregated
# bias_agg <- aggregate(bias1, fact=50,FUN=mean)
# writeRaster(bias_agg, filename="bias_agg.asc", datatype='ascii',
#            overwrite=TRUE,NAflag=-9999)

## since I used setwd() to the output folder, it is automatically stored there,
## if not, paste the path to the filename, i.e. 
## filename=paste0(output_wd,"/bias_2023.asc") #or # here(output_wd,"bias_2023.asc")
terra::writeRaster(bias1, filename="bias_2025.asc",
            overwrite=TRUE,NAflag=-9999)
```

Now, create folders for the MaxEnt output by hand or do it with the following code. We want a subfolder in our output-folder, called MaxEntRes, with two subfolders called 'bias' and 'nobias'.

```{r}
dir.create(path = file.path(output_wd, "MaxEntRes"), showWarnings = FALSE)
dir.create(path = file.path(output_wd, "/MaxEntRes", "nobias_current"),showWarnings = FALSE)
dir.create(path = file.path(output_wd, "/MaxEntRes", "nobias_future"),showWarnings = FALSE)
dir.create(path = file.path(output_wd, "/MaxEntRes", "withbias_current"),showWarnings = FALSE)
dir.create(path = file.path(output_wd, "/MaxEntRes", "withbias_future"),showWarnings = FALSE)
```

Before we work with MaxEnt, please inspect the header of your bias file by opening it in a normal text editor. Mine looks like: <br> <br> 

NCOLS 1423 <br> 
NROWS 1425 <br> 
XLLCORNER 108.334122887 <br> 
YLLCORNER -4.374012999 <br> 
CELLSIZE 0.0083333337997189 <br> 
NODATA_value -9999 <br> 

If this throws an error in MaxEnt, please change the <br>
CELLSIZE from 0.0083333337997189 to **0.0083333338** by hand! That is, the header should read: <br>


ncols         1423   <br> 
nrows         1425   <br> 
xllcorner     108.33412288693  <br> 
yllcorner     -4.3740129986359 <br> 
cellsize      0.0083333338  <br> 
NODATA_value  -9999   <br> 

(Please, without '<br>' if you use copy-paste from my .rmd course script.....)

+++++++++++++++++++++++++++++++++++++++++++++++++
Exercise: Please run MaxEnt manually first by opening the maxent.jar 
(double click -> then the grey GUI should open after some seconds) - see lecture.
This helps you to understand the principles first before running the code in R.
+++++++++++++++++++++++++++++++++++++++++++++++++

# Running MaxEnt in R


## select predictors
Here you can select which predictors you want to use. Please remember from the
above exercise that current and future maps need to carry the same name,
but in different folders, and that you must have the future
maps for those variables that you included to fit the model to current conditions! 
That means, if you have fitted the model to current temperatures, you need
to have  maps of future temperatures as well (with the same name(s)) 
if you want to project
the model to future climates.

```{r}
files # files <- list.files(path= maps_wd, pattern='.asc$',full.names=TRUE )

## Let's select the first 6 to eliminate the hillshade
predictors_selected <- terra::rast(x = c(files[c(1:6)]) ) 
```


Predictors are then stacked in the following MaxEnt command. 
Also, don't forget to mark categorical variables and to not select
variables which do not contain biological information, like the hillshade.

## Maxent without bias correction
Here, the whole area that is given by the raster is sampled equally (background).


### MaxEnt without bias current

```{r}
xm_nobias_current <- dismo::maxent(
                            x = raster::stack(predictors_selected), # convert to raster stack, does not work with terra
                            p = sp_recs[,2:3], # x and y coordinates
                            factors = 'bio_asc_42', # raster names which are factors
                            nbg = nrow(sp_recs) * 10, # number of background points, ten times
                            path = file.path(output_wd, "/MaxEntRes", "nobias_current"), # output path
                            args = c("replicates=3", 
                                     "outputformat=logistic", 
                                     "responsecurves") # number of replicates and output format
                            )

saveRDS(xm_nobias_current, file.path(output_wd, "/MaxEntRes", "nobias_current", "maxent_nobias_current.rds"))
```

### predict without bias current

```{r}
pred_no_bias_current <- terra::mean(raster::predict(xm_nobias_current, raster::stack(predictors_selected)))

writeRaster(pred_no_bias_current, 
            file.path(output_wd, "/MaxEntRes", "nobias_current", "species_selection_cur_avg.asc"),
            overwrite = TRUE)
```

### MaxEnt without bias future

```{r}
xm_nobias_future <- dismo::maxent(
                            x = raster::stack(predictors_selected), # convert to raster stack, does not work with terra
                            p = sp_recs[,2:3], # x and y coordinates
                            factors = 'bio_asc_42', # raster names which are factors
                            nbg = nrow(sp_recs) * 10, # number of background points, ten times
                            path = file.path(output_wd, "/MaxEntRes", "nobias_future"), # output path
                            args = c("replicates=3", 
                                     "outputformat=logistic", 
                                     "responsecurves",
                            paste0("projectionlayers=", maps_wd_fut)) 
                            )

saveRDS(xm_nobias_future, file.path(output_wd, "/MaxEntRes", "nobias_future", "maxent_nobias_future.rds"))
```

## MaxEnt with sampling bias correction
### Create background points

```{r}
random_pts <-  dismo::randomPoints(mask = raster(bias1),
                                   n = nrow(sp_recs) * 10 ,
                                   prob = T)


```

### MaxEnt with bias current

```{r}
xm_withbias_current <- dismo::maxent(
                            x = raster::stack(predictors_selected), # convert to raster stack, does not work with terra
                            p = sp_recs[,2:3], # x and y coordinates
                            factors = 'bio_asc_42', # raster names which are factors
                            a = random_pts, # bias - background points
                            path = file.path(output_wd, "/MaxEntRes", "withbias_current"), # output path
                            args = c("replicates=3", 
                                     "outputformat=logistic", 
                                     "responsecurves") # number of replicates and outputformat
                            )

saveRDS(xm_withbias_current, file.path(output_wd, "/MaxEntRes", "withbias_current", "maxent_withbias_current.rds"))
```

### predict with bias current

```{r}
pred_with_bias_current <- terra::mean(raster::predict(xm_withbias_current, raster::stack(predictors_selected)))

writeRaster(pred_with_bias_current, 
            file.path(output_wd, "/MaxEntRes", "withbias_current", "species_selection_cur_avg.asc"),
            overwrite = TRUE)
```

### MaxEnt with bias future

```{r}
xm_withbias_future <- dismo::maxent(
                            x = raster::stack(predictors_selected), # convert to raster stack, does not work with terra
                            p = sp_recs[,2:3], # x and y coordinates
                            factors = 'bio_asc_42', # raster names which are factors
                            a = random_pts, # bias - background points
                            path = file.path(output_wd, "/MaxEntRes", "withbias_future"), # output path
                            args = c("replicates=3", 
                                     "outputformat=logistic", 
                                     "responsecurves",
                            paste0("projectionlayers=", maps_wd_fut))
                            )

saveRDS(xm_withbias_future, file.path(output_wd, "/MaxEntRes", "withbias_future", "maxent_withbias_future.rds"))
```


# Accessing MaxEnt results

## List the results

```{r}
# Have a look at the MaxEnt layers produced
# MaxEnt saves the mean of x repetitions in _avg.asc

# The script now only works with the folder structure
# MaxEntRes/nobias
# MaxEntres/withbias

# new argument 'recursive' means, that also subfolders are checked!
infiles <- list.files(path=paste(output_wd,'/MaxEntRes',
                                 sep=''),pattern='_avg.asc$',
                      full.names=TRUE,recursive=TRUE )
infiles


# example
#[1] "c:/Users/kramer/Dropbox/.../MaxEntRes
#     /nobias/river_dummy_avg.asc"
#[2] "c:/Users/kramer/Dropbox/.../MaxEntRes
#     /nobias/river_dummy_FutureMaps_avg.asc"
#[3] "c:/Users/kramer/Dropbox/.../MaxEntRes
#     /withbias/river_dummy_avg.asc"
#[4] "c:/Users/kramer/Dropbox/.../MaxEntRes
#     /withbias/river_dummy_FutureMaps_avg.asc"
```

## Stack results file and make usual plot
```{r}
## IF you have a slow computer, 
## please use the workaround with function aggregate() from above:
# me_stack <- aggregate(c(infiles[c(1:4)]),fact= 50, FUN = mean )

me_stack <- c(rast(infiles[[1]]),
              rast(infiles[[2]]),
              rast(infiles[[3]]),
              rast(infiles[[4]]))

# name sequence according to infiles list above
names(me_stack) <- c('curr_noBias','fut_noBias',
                     'curr_withBias','fut_withBias')

#par(mar=c(4,4,3,3), oma=c(2,2,2,2))

plot(me_stack,col=viridis(100)) # note: I might use a different example than you did in MaxEnt
```

Check the range of the values

```{r}
boxplot(me_stack,layers = c(1,3,2,4),notch=T,outline=F)
```

How would you interpret the results?

## Threshold selection

You should have results for the runs with a bias file and without:
```{r}
# Use maxentResults.csv to access threshold value
# column: 10th percentile training presence logistic threshold

me_res <- list.files(path=paste(output_wd,'/MaxEntRes',sep=''),
                     pattern='maxentResults.csv',
                     full.names=TRUE,recursive=TRUE )
me_res

# example
#[1] "c:/Users/kramer/Dropbox/.../MaxEntRes
#     /nobias/maxentResults.csv"
#[2] "c:/Users/kramer/Dropbox/.../MaxEntRes
#     /withbias/maxentResults.csv"
```

Now we do some elegant trick: we store both files as two objects in a list! 
Similar to stacking rasters, a bit....
```{r}
store_res <- lapply(me_res,FUN=read.csv) #store as list obj.
## Check the following out:
# head(store_res) # list object with 2 slots
# str(store_res)
# store_res[[1]]
```

We are searching for the column with the threshold.
```{r}
## Extract the values at column '10th percentile' at position nrep +1
## (= nrow command) to access avg value of all repetitions at the lowest row

zename<-'X10.percentile.training.presence.Logistic.threshold' # check if the column name is the same !!!!!
## Achtung! 
## In MaxEnt 3.4.1 the column name is:
## 'X10.percentile.training.presence.Logistic.threshold'
## with capital L in logistic
zename
```

Now check which column number that is....

```{r}
zecol <-which(colnames(store_res[[1]]) == zename) 
zecol
```

... and extract the threshold value :
```{r}
## a little head twister:
## access each element of your list, which is a whole dataset with store_res[[1]]
## and inside this list object = data.frame, access the element of
## the last row (nrow), and the column zecol] containing the threshold value
t_noBias   <- store_res[[1]][nrow(store_res[[1]]),zecol]
t_withBias <- store_res[[2]][nrow(store_res[[2]]),zecol]

t_noBias #the thresholds
t_withBias
```

Now apply the thresholds to your MaxEnt output maps of relative probability. For that, store the thresholds in a vector, and double them for bias/ nobias, because the threshold stays the same:

```{r}
all_bias <- rep(c(t_noBias,t_withBias), each=2)
all_bias
```

Now set the thresholds to the stack. If you don't remember the stack (me_stack), plot it again. Test different thresholds and see how the predictions change:

```{r}
binary_thresh <- me_stack >= all_bias
plot(binary_thresh)
```

```{r}
test05 <- me_stack >= 0.5 # check result with fixed threshold of default cut-off
plot(test05)

#testSeq <- me_stack >= seq(0.1, 0.7, length = 4)
#plot(testSeq)
```

## Analyses

Now let's check how global change affects the protected areas. Does the suitability of these areas increase or decrease in the future?
```{r}
# Extract relative probability values inside protected areas PAs
# check if overlay correct - only works when you have NOT aggregated the maps
terra::compareGeom(PA_ras,me_stack[[1]]) 

ex <- which(values(PA_ras)==1) ##remember PA_ras == 1 where the PAs are
# ex # gives Pointer to elements/ Index
ex_stack <- terra::extract(x=me_stack, y=ex)
head(ex_stack)
```

Plot the difference of habitat suitability in scenarios and PAs
```{r}
boxplot(ex_stack,na.rm=T)
```

Calculate no. of cells (=area) of suitable habitat in PAs

```{r}
ex_binary <- terra::extract(x=binary_thresh, y=ex)
head(ex_binary)
## In the following, we are summing the cells with 'TRUE'
colSums(ex_binary,na.rm=T) 
```

Interestingly: when correcting for sampling bias, the future scenarios have the lowest number of suitable habitat inside PAs: <br>
<br>
curr_noBias    fut_noBias curr_withBias  fut_withBias   <br>
      43285         63905         43430         39989   <br>
      
      
```{r, echo=FALSE, eval=FALSE}
################# just in case for Netlogo   ###################################
########## maps have to be smaller and instead of NA put -1 ####################
################################################################################

infiles # remember, output folder content

# chose one:
my_MaxEnt_avg_Map <- terra::rast(infiles[1])
# my_MaxEnt_avg_Map@crs <- CRS("+proj=longlat +datum=WGS84") #not needed with {terra}-package

# plot
plot(my_MaxEnt_avg_Map,col=viridis(100))

# save it with a lower resolution (similar to 'aggregate')
+probab_agg <- terra::aggregate(x=my_MaxEnt_avg_Map,fact=2,fun=mean)

#and make an integer out of the decimals
probab_agg_int <- round(probab_agg*100, digits=0)
plot(probab_agg_int, col=viridis(100))

# save it to import into Netlogo
getwd() # that's where it is saved now
terra::writeRaster(probab_agg_int, filename="probab_agg_int.asc",
                   datatype="ascii",
                   overwrite=TRUE,NAflag = -9999)
```

# Least cost path for connectivity

```{r}
#install.packages('gdistance')
library(gdistance)
#?gdistance
# we need a workaround, because we need the PAs now as SpatialPolygonDataFrame, 
# and not as sf-Object (gdistance works with old format)

#PA_shp_sp <- as(PA_shp_sf, "Spatial")

# Set start and end points: centers of the first two PA polygons
start <- st_centroid(PA_shp_sf[1,1])
end   <- st_centroid(PA_shp_sf[2,1])

# Several commands in one line when separated by ';':
plot(me_stack[[1]]); points(start, col = "red"); points(end, col = "orange")

# Clip raster to save computation time
cr_extent <- c(114,117,3,6.5)
me_cr <- crop(x=me_stack[[1]],y=cr_extent)
plot(me_cr); points(start,pch=15, col = "red"); points(end,pch=15, col = "orange")

### Necessary calculations
# Calculate the transition layer

# but first, we need to backtransform the SpatRaster {terra}-Object
# into a RasterLayer {raster}-Object:
me_cr_raster <- raster::raster(me_cr)

trans <- gdistance::transition(x = me_cr_raster,
                    transitionFunction = mean,
                    directions = 4,
                    symm = F)
class(trans)

# Calculate the shortest weighted connection
sPath_sp <- gdistance::shortestPath(trans, as(start, "Spatial"), as(end, "Spatial"),
                                    output="SpatialLines")

terra::crs(sPath_sp) <- "+proj=longlat +datum=WGS84 +no_defs"
# not needed any more
# crs(sPath_sp) <- "+proj=longlat +datum=WGS84 +no_defs"


# Calculate the length of the path
gdistance::costDistance(trans, as(start, "Spatial"), as(end, "Spatial")) #units?


# Make a plot
plot(me_cr,col=viridis(100)); points(start,pch=17,col='white'); points(end,pch=15, col='white')
lines(sPath_sp,lwd=1, col= 'red'); plot(PA_shp_sf, border='white', lwd=0.5,add=T)
```


```{r, echo=FALSE, eval=FALSE}
# where does the animal corridor intersect rivers?
# in sf package
River_inter1 <- sf::st_intersection(x = River_shp_sf, y = as(sPath_sp, "sf")) 
# in sp package
# River_inter_sp <- gIntersection(River_shp, sPath) 

plot(sPath_sp,lwd=3); plot(PA_shp_sp,border='grey',lwd=1,add=T)
plot(River_shp_sf,col='blue',lwd=2,add=T)
plot(River_inter1,col='red',lwd=4,add=T); box()
```


```{r, echo=FALSE, eval=FALSE}
## Which connections intersect the PAs?

## as PA_shp_sf has an invalid topology, we need the workaround with the old 
## {rgeos}package; https://github.com/r-spatial/sf/issues/1902
# PA_inter1 <- st_intersection(x = PA_shp_sf, y = as(sPath_sp, "sf")) 
PA_inter1 <- sf::st_intersection(sf::st_make_valid(PA_shp_sf), as(sPath_sp, "sf")) # old way

plot(sPath_sp,lwd=3); plot(PA_shp_sf,border='blue',lwd=2,add=T)
lines(PA_inter1,col='red',lwd=4); box()
```


```{r, echo=FALSE, eval=FALSE}
### Save new line as shapefile
cp_line <- as(sPath_sp, "sf")
cp_line$name <- c(1:length(sPath_sp@lines))

sf::st_write(obj = cp_line, dsn = paste0(output_wd,'/costpath_line.shp'), delete_layer = TRUE)

## TODO: check out this package:
# https://cran.r-project.org/web/packages/leastcostpath/vignettes/leastcostpath-1.html

################  end #######################
```

##### species data

```{r}
# data_spec <- read.delim(here("data", "data_borneo", "animal_data", "old", "allSpecies20092011.txt"))
# 
# data_spec$Spec_lat |> table()
# 
# data_sub <- data_spec %>% 
#   dplyr::select(Spec_lat, GPS_Long, GPS_Lat) %>% 
#   filter(Spec_lat %in% "Pardofelis marmorata") |> 
#   dplyr::rename(species = "Spec_lat",
#                 long = "GPS_Long",
#                 lat = "GPS_Lat") |> 
#   filter(!long %in% 0)
# 
# write.csv(data_sub, here("data", "data_borneo", "animal_data", "Pardofelis_marmorata.csv"), row.names = FALSE)
```````


