---
title: "Analysis of biodiversity data"
author: "Aimara Planillo"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y')) `"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: show
    toc_depth: 4
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}
p {
  line-height:170%;
}
#sidebar h2 {
  background-color: Indigo;
}
code {
  color: Indigo ;
}
.exercise {
  color: #824b00;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      dev = "ragg_png", fig.width = 9, fig.height = 6, dpi = 600, retina = 1)
Sys.setlocale("LC_TIME", "C")
```


# Setup


```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', fig.width = 6, fig.height = 6)
library(knitr)
library(vegan)
library(iNEXT)
library(sads)
library(tidyverse)
library(terra)
library(here)
```


# SAD curves

We can obtain the Species Abundance Distribution for our samples
To use the library `{sads}`, and the example data on moth abundance.

```{r}
data(moths)
moths
?moths
```

## Preston plot
First, we are going to obtain the Preston plot, also known as octaves. 
This plot represents the number of species in classes of logarithm of abundances at base 2.

```{r octaves}
moths.oc <- octav(moths)
moths.oc
plot(moths.oc)
```


## Whittaker diagram
We can also obtain the Whittaker plot for our data.
This is the rank-abundance diagram
```{r RAD}
# for the moths data
moths.rad <- rad(moths)
plot(moths.rad, ylab="Number of individuals")
```


## Species Abundance Distributions: fitting curves

We are going to fit some curves to the rank abundance plot

```{r}
# build the model
moths.ge <- fitsad(x=moths, sad="geom") # geometric distribution
moths.ls <- fitsad(x=moths, sad="ls") # log series distribution
moths.ln <- fitsad(x=moths, sad="lnorm") #log-normal distribution

# get rank abundance objects
moths.ge.rad <- radpred(moths.ge)
moths.ls.rad <- radpred(moths.ls)
moths.ln.rad <- radpred(moths.ln)


# Plot the curves
plot(moths.ln.rad)
plot(moths.ln.rad, xlab = "Rank", ylab = "Abundance", log = "y",
     type = "l", col = "green", lty = 1, lwd = 6)

# We can superimpose the curve to the rank plot
plot(moths.rad)
lines(moths.ge.rad, col="red")
lines(moths.ls.rad, col="blue")
lines(moths.ln.rad, col="green")
legend("topright",c("Geometric", "Logseries", "lognormal"),lty=1, col=c("red", "blue", "green"))

## looking at the fits
logLik(moths.ge)
logLik(moths.ls)
logLik(moths.ln)
```



###########################################
# Alpha diversity
###########################################


## Shannon index

Let us calculate Shannon index using example data from the power point

```{r shannon_birds}
birds1 <- data.frame(Species = c('BlueTit', 'Robin', 'Magpie',
                                     'GreatTit'),
                         Abundance = rep(9, 4))
birds1


## now let us get the pi, ln(pi), N and S to calculate Shannon index
N <- sum(birds1$Abundance)
S <- nrow(birds1)

pi <- birds1$Abundance/N
lnpi <- log(pi)

H <- -sum(pi*lnpi)
H
```

## Community structure: `{vegan}` package

vegdist() function allows calculating multiple community dissimilarity indices.

```{r shannon_vegan}
# 1. transpose the data
birds1.transpose <- as.data.frame(t(birds1[, -1]))
colnames(birds1.transpose) <- birds1$Species
birds1.transpose

# Get diversity value
?diversity
H_vegan <- diversity(birds1.transpose, index = "shannon")
H_vegan
```


## Properties of Diversity indices
Shannon diversity vs Hill numbers

If we have two communities with no species in common, the diversity of both together should be the sum of the diversities of each one
```{r shannon_vs_hill}
# We create a second community with no species in common with the first one
birds2 <- data.frame(Species = c('Sparrow', 'Dove', 'Crow'),
                         Abundance = c(4,5,20))
birds2 

# the transpose matrix for the analysis
birds2.transpose <- as.data.frame(t(birds2[, -1]))
colnames(birds2.transpose) <- birds2$Species
birds2.transpose


# Both communities in the same table 
# transpose data and get sums 

birds.both <- merge(birds1, birds2, by = 'Species', all = T)
birds.both$Abundance.x[is.na(birds.both$Abundance.x)] <- 0
birds.both$Abundance.y[is.na(birds.both$Abundance.y)] <- 0
birds.both
birds.all <- rowSums(birds.both[,2:3])
birds.all

both.trans <- as.data.frame(t(birds.both[, -1]))

colnames(both.trans) <- birds.both$Species
rownames(both.trans) <- c("birds1", "birds2")
both.trans

all.trans <- colSums(both.trans)

# shannon diversity for each sample and for the sum
H1 <- diversity(both.trans, index = "shannon")
H2 <- diversity(all.trans, index = "shannon")

H1
H2


# Hill number order 1 (library iNEXT)
HN1.birds1 <- iNEXT(birds1$Abundance)
HN1.birds1$AsyEst

HN1.birds2 <- iNEXT(birds2$Abundance)
HN1.birds2$AsyEst

HN1.birdsboth <- iNEXT(birds.all)
HN1.birdsboth$AsyEst
```



# Hill Numbers

## Get the **Hill Number** with q = 0 (Species richness)

Load data from the library about spider samples in two locations
```{r load_inext_data}
data(spider)
str(spider)
?spider
```

```{r run_inext_q0}
example1 <- iNEXT(spider, q = 0, datatype = "abundance")

example1$DataInfo
#Show a summary of the data with diversity estimates in rarefied and extrapolated samples
example1$iNextEst
# show asymptotic estimates
example1$AsyEst
```


## Rarefaction/extrapolation

### Get new values

We can specify the **sample size** for which we want to **rarefy/extrapolate**
For Species richness, the *extrapolation* is reliable up to *double* the reference sample. 
For q = 1 or 2, the extrapolation can be extended to the asymptote.

We run the code simultaneously for the three Hill numbers:

```{r}
# We define the number of samples size that we want to use for estimation
m <- c(1, 50, 100, 200, 400)
example2 <- iNEXT(spider, q = c(0,1,2), datatype = "abundance", size = m)
example2$iNextEst 
```


### Plotting

Plot for an iNEXT object created previously
Rarefaction/extrapolation curves for q = 0
```{r}
ggiNEXT(example1, type=1) # Curve for sample size
```

Rarefaction/extrapolation curves for sample size for all 3 numbers divided by site
```{r}
ggiNEXT(example2, type=1, facet.var="Assemblage")
```

Rarefaction/extrapolation curves for sample size separated by "order" (q)
```{r}
ggiNEXT(example2, type=1, facet.var="Both")
?ggiNEXT
ggiNEXT(example2, type=1, facet.var="Order.q")
# adding grey=TRUE, we get a plot in black and white theme
```

We can also work with **sample coverage**

```{r}
ggiNEXT(example1, type=3) # Curve for sample size
```


## Define at which sample size you want to compare your samples. 

Results: for the three hill numbers, order 0, 1,2. 
For all of them is interpolated or extrapolated as requested and we get the values in qD with lower and upper confidence intervals. 
These values are now fully comparable

Use max value of the biggest sample size

```{r}
inext_spiders <- iNEXT(spider, q = 0, datatype = "abundance")
info_spiders <- inext_spiders$DataInfo
max(info_spiders$n)

estINEXTsize <- estimateD(spider, datatype = "abundance", base = "size", level = 252,
          conf = 0.95)
estINEXTsize
```


Use max sample coverage 

```{r}
max(info_spiders$SC)

estINEXTcov <- estimateD(spider, datatype = "abundance", base = "coverage", level = .945,
          conf = 0.95)
estINEXTcov
```


Now we can compare Species richness standardized for the same level of sample size for each habitat (sample)

```{r}
estINEXTcover2 <- estimateD(spider, datatype = "abundance", base = "size", level = 252,
          conf = 0.95) 

habitat <- factor(c("Girdled", "Logged"))

# plot simple
mysub <- subset(estINEXTcover2, estINEXTcover2$Order.q == 0 )
plot(mysub$qD ~ habitat, border = c("green4", "red"), 
        xlab = "Habitats", ylab = "Est. Species richness")

# plot with confidence interval in ggplot
ggplot(mysub, aes(x = habitat, y = qD)) +
  geom_point(colour = c("green4", "red"), size = 5) +
  geom_errorbar(aes(ymin = qD.LCL, ymax = qD.UCL), 
                colour = c("green4", "red"),
                width = 0.2) +
  xlab("Habitats") +
  ylab("Est. Species richness") +
  theme_bw() 
```



##########################################
# SPATIAL ANALYSIS OF DIVERSITY
##########################################

## BIRD DATA

How does urbanization affect species richness?


Load data
```{r}
# Bird data
bird_data <- read.csv(here("data","data_berlin","animal_data",
                           "Birds_Berlin_exercise_planillo2021.csv") )
head(bird_data)
str(bird_data)
summary(bird_data)
bird_data
```

## Get the **Hill Number** for species richness

iNext package uses the data in a specific format:
Matrix with species in rows, sites in columns.

```{r run_inext_birds}
# get the data in the proper format
bird_data <- column_to_rownames(bird_data, "site") 
bird_data <- t(bird_data)

# run inext function
birds_inext <- iNEXT(bird_data, q = 0, datatype = "abundance") # q = 0 -> species richness

#Show a summary of the data 
birds_inext$DataInfo
#Show a summary of the data with diversity estimates in rarefied and extrapolated samples
head(birds_inext$iNextEst)
# show asymptotic estimates
birds_inext$AsyEst

### a little workaround to create 29 colours and dot types ####
library(scico)
library(paletteer)
paletteer::paletteer_c("scico::berlin", 29)
my_palette_inext <- paletteer::paletteer_c("scico::berlin", 29)
# Species accumulation curves
ggiNEXT(birds_inext, type=1, facet.var="None") + # not all plots sampled equally
scale_color_manual(values = my_palette_inext) +
  scale_fill_manual(values = my_palette_inext) +
  scale_shape_manual(values = seq(1:29))


# get minimum number of individuals from data 
min_abund <- min(birds_inext$DataInfo$n)
bird_data2 <- as.data.frame(bird_data)
str(bird_data2)

# use 2x minimum number of individuals for rarefaction/extrapolation
birds_estINEXTsize <- estimateD(bird_data, q = 0, datatype = "abundance", base = "size", level = (min_abund * 2),
          conf = 0.95)

birds_estINEXTsize

# extract species richness 
birds_est_sprich <- as.data.frame(cbind(site = colnames(bird_data),
                                        sp_rich = birds_estINEXTsize$qD))

birds_est_sprich
```


# Statistics!!

# Question: Does urbanization have an effect on bird diversity?

We are going to run a model with the number of species as response and 
urbanization variables as explanatory variables.

## Load environmental data
```{r}
env_cov <- read.csv(here("data","data_berlin","animal_data",
                         "birds_transects_allenvir_100m.csv") )
head(env_cov)
str(env_cov)
summary(env_cov)

```

We are going to use three variables to define the ubanization gradient:
tree cover, imperviousness and noise.

## Prepare the data for the model 
```{r}
colnames(env_cov)
# Put all data together: add environmental variables to birds data
#my_model_data <- left_join(birds_est_sprich, env_cov, by = "site")
my_model_data <- merge(birds_est_sprich,env_cov,by = "site")

#select response and explanatory variables
#my_model_data <- dplyr::select(my_model_data, 
#                               c(site, sp_rich, tree_cover, impervious_surface, noise))

my_model_data <- my_model_data[, c('site', 'sp_rich', 'tree_cover', 
                                   'impervious_surface', 'noise')]

str(my_model_data) # do you also have a chr for sp_rich?
my_model_data$sp_rich <- as.numeric(my_model_data$sp_rich)
str(my_model_data)

```

## Define the model
```{r}
# explore relationships between variables
ggplot(my_model_data, aes(y = sp_rich, x = tree_cover)) +
  geom_point() +
  geom_smooth()

ggplot(my_model_data, aes(y = sp_rich, x = impervious_surface)) +
  geom_point() +
  geom_smooth()

ggplot(my_model_data, aes(y = sp_rich, x = noise)) +
  geom_point() +
  geom_smooth()

# build linear regression model
birds_model <- glm(sp_rich ~ tree_cover + impervious_surface + noise,
                    family = "gaussian", 
                    data = my_model_data)

# View results of the model
summary(birds_model)
plot(birds_model)

# bird diversity decreases with increasing urbanisation


# view the regression line through impervious surface only:
ggplot(my_model_data, aes(y = sp_rich, x = impervious_surface)) +
  geom_point(size=7,alpha=0.5) +
  geom_smooth(method = "lm", se = TRUE, col='red') +
  xlab('impervious surface')

```


# **PREDICT** the model in Berlin

## Load environmental rasters
```{r}
##
## set working directory for maps, e.g. here geoTiffs ##
tree_raster   <- rast(here::here("data","data_berlin","geo_raster_current_gtif","tree_cover_density_2012_100m_3035.tif"))
imperv_raster <- rast(here::here("data","data_berlin","geo_raster_current_gtif","imperviousness_2012_100m_3035.tif"))
noise_raster  <- rast(here::here("data","data_berlin","geo_raster_current_gtif","noise_daynight_2017_100m_3035.tif"))
water_raster  <- rast(here::here("data","data_berlin","geo_raster_current_gtif","water_bodies_2010_100m_3035.tif"))

#put all environmental rasters together for the prediction
#many_rasters <- list(x,x)

## this works
my_env_stack <- rast(list(tree_raster, imperv_raster, noise_raster))

# the raster the same name as the variables in the model
names(my_env_stack) <- c("tree_cover", "impervious_surface", "noise")


# the model is fitted with environmental predictor values between 0 and 100
# e.g. tree cover = 88.5 % in a 100*100 m cell
# check: my_model_data
# However, our rasters do not contain decimals (for PC storage and memory reasons)
# so we need to transform them to decimals before we predict our model
# to the whole of Berlin:

my_env_stack_2 <- my_env_stack/100 # correct values of the rasters
```

## Predict the model
```{r}
sp_rich_pred <- terra::predict(object = my_env_stack_2, 
                                model = birds_model)
sp_rich_pred
sp_rich_pred[sp_rich_pred < 0] <- 0 # Abundance cannot be < 0

# define colors
my_palette <- c("#440154FF", "#2D708EFF", "#56C667FF", "#DCE318FF", "#FDE725FF")

# plot map
plot(sp_rich_pred, col = my_palette, breaks = c(seq(5, 55, by = 10)))
plot(water_raster, col = "darkslategray1",  legend=FALSE, add = TRUE)

# writeRaster(sp_rich_pred, my_output_directory)
```

##########################################

### EXTRA INFO

##########################################

In iNEXT package also calculates asymptotic diversity metrics. 
The estimated asymptotes area calculated via the functions

* `ChaoRichness()` for q = 0
* `ChaoShannon()` for q = 1
* `EstSimpson()` for q = 2

For example, to estimate the species richness
```{r}
ChaoRichness(spider, datatype = "abundance")
```



# <span class='exercise'>Exercise</span>

Some exercises.
```{r}
###
## Diversity Analysis Exercise

# 1. Load the data
## - 1.1. the bird data 'birds_berlin_exercise_planillo2021.csv' and explore it (use head(), str())
## - 1.2. the environmental data 'birds_transects_allenvir_100m.csv' and explore it  

# 2. Estimate alpha diversity:
## - 2.1. Choose the diversity index: q = 1 or q = 2
## - 2.2. Rarefy the samples to the appropriate size and estimate the rarefied values

# 3. Predict Hill number 1 or 2 in Berlin
## - 3.1. Choose environmental variables (up to 3) 
## - 3.2. Run model with the diversity values and the environmental variables
## - 3.3. Plot the predictions

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~#
#### for those who like challenges
# 4. Estimate beta diversity:
## 4.1. Jaccard dissimilarity (for each pair of sites)
## 4.2. Bray-Curtis dissimilarity (for each pair of sites) 
## 4.3. Compare the output of both beta-diversity metrics

```


<br><hr><br>

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
#git2r::repository() ## uncomment if you are using GitHub
sessionInfo()
```

</details>
