---
title: "Analysis of biodiversity data"
author: "Aimara Planillo, Viktoriia Radchuk"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y')) `"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: show
    toc_depth: 4
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}
p {
  line-height:170%;
}
#sidebar h2 {
  background-color: Indigo;
}
code {
  color: Indigo ;
}
.exercise {
  color: #824b00;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      dev = "ragg_png", fig.width = 9, fig.height = 6, dpi = 600, retina = 1)
Sys.setlocale("LC_TIME", "C")
```

# Setup

```{r}
library(knitr)
library(vegan)
library(psych)
library(here)
library(ape)
```


# Beta diversity

## 1. Jaccard

Let's calculate Jaccard diversity using the data from the power point

```{r Jaccard_birds}

# Create dataframes for the two communities:
Gruenewald <- data.frame(Species = c('BT', 'Robin', 'Magpie','Roller', 'Swallow', 'GT'),
                         Abundance = c(3, 2, 2, 1, 2, 4))
Gruenewald

Alex <- data.frame(Species = c('BT', 'Dove', 'Magpie', 'GT', 'Crow'),
                   Abundance =  c(3, 4, 1, 5, 2))
Alex

# Now let's get a, b and c needed to calculate Jaccard index

Common <- merge(Gruenewald, Alex, by = 'Species') # create a dataframe with only the common species of the two communities
a <- nrow(Common) # get the number of common species
Unique_Grun <- sum(! Gruenewald$Species %in% Alex$Species) # get the number of unique species for Grunewald
Unique_Alex <- sum(! Alex$Species %in% Gruenewald$Species) # get the number of unique species for Grunewald
J <- a / (a + Unique_Grun + Unique_Alex) # calculate Jaccard index
J 

```


## 2. Bray-Curtis

```{r BC_birds}

# Create dataframes for the two communities:
Gruenewald <- data.frame(Species = c('BT', 'Robin', 'Magpie','Roller', 'Swallow', 'GT'),
                         Abundance = c(3, 2, 2, 1, 2, 4))
Gruenewald

Alex <- data.frame(Species = c('BT', 'Dove', 'Magpie', 'GT', 'Crow'),
                   Abundance =  c(3, 4, 1, 5, 2))
Alex

# Now let's get a, b and c needed to calculate Jaccard index

Common <- merge(Gruenewald, Alex, by = 'Species') # create a dataframe with only the common species of the two communities
Common$Min <- unlist(lapply(1:nrow(Common), FUN = function(x){min(Common[x, 2:3])})) # get the minimum abundance of each common species between the two communities
Common

S_Grun <- sum(Gruenewald$Abundance) # get the total abundance of all species for Grunewald
S_Alex <- sum(Alex$Abundance) # get the total abundance of all species for Alex
BC <- 2*sum(Common$Min) / (S_Grun + S_Alex) # calculate Bray-Curtis 
BC

```


# Principal Coordinates Analysis  

## Bird data  

First, let's read the .csv data

```{r PCoA_readData}

# Export dataframe of species abundance and environmental variables for each site
species <- read.csv(here('data','data_berlin','animal_data','birds_berlin_exercise_planillo2021.csv'))
envir   <- read.csv(here('data','data_berlin','animal_data','birds_transects_allenvir_100m.csv'))

# Let's look how the bird species data look like
head(species) 
str(species) #29 transects
# Data for great tits (Parus major)
hist(species$Parus_major, col = 'grey', main = '', xlab = 'Great tit abundance') 

# Let's look how the env. data looks like
head(envir) # every line/ row corresponds with a site
hist(envir$pop_100m, col = 'grey', main = '', xlab = 'Human population density')
hist(envir$noise, col = 'grey', main = '', xlab = 'Noise level')
```


And now let's fit a PCoA on the species data
```{r PCoA_species}
# remove site as a column
bird_data <- species[,-71]
rownames(bird_data) <- species$site

# Bray curtis distance between sites based on species abundances
bird_dist <- vegdist(bird_data, method = "bray")

# Principal coordinates analysis or metric multidimensional scaling
birds_pcoa <- cmdscale(bird_dist, k = nrow(bird_data)-1, list. = TRUE)

# Plot of the sites
ordiplot(scores(birds_pcoa, choices=c(1,2)), type="t", main="PCoA with species weighted averages")
abline(h=0, lty=3)
abline(v=0, lty=3)

scores(birds_pcoa)

# Add weighted average projection of species to plot
birds_wa <- wascores(birds_pcoa$points[,1:2], bird_data)
text(birds_wa, rownames(birds_wa), cex=0.7, col="red")

```



```{r PCoA_species2}
## Plot with species weights

# Add weighted average projection of species
#birds_wa <- wascores(birds_pcoa$points[,1:2], bird_data)
#text(birds_wa, rownames(birds_wa), cex=0.7, col="red")
```

## Add environmental values

```{r PCoA_species3}
# Add environmental variables
rownames(envir) <- (envir$site)
envir_data <- envir[,-1]

# population density
pop_fit <- envfit(birds_pcoa$points, envir_data$pop_100m)

ordiplot(scores(birds_pcoa, choices=c(1,2)), type="t", main="PCoA with species weighted averages")
abline(h=0, lty=3)
abline(v=0, lty=3)
plot(pop_fit, p.max = 0.05, col = "red")

# multiple environmental variables
head(envir_data)
envir_data2 <- envir_data[,c("impervious_surface", "pop_100m", "noise", "tree_cover", "open_green")]

env_fit <- envfit(birds_pcoa$points, envir_data2)

ordiplot(scores(birds_pcoa, choices=c(1,2)), type="t", main="PCoA with species weighted averages", 
         cex = 1)
abline(h=0, lty=3, col = "grey80")
abline(v=0, lty=3, col = "grey80")
# points(birds_wa, col = "blue")
text(birds_wa, rownames(birds_wa), cex=0.7, col="blue")
plot(env_fit, col = "red", bg = "grey80")
```


## Overlap the grouping of the sites

```{r PCoA_species4}
# grouping the sites
envir$imperv_classes <- ifelse(envir$impervious_surface < 25, "low", 
                              ifelse(envir$impervious_surface > 75, "high", "medium"))


plot(birds_pcoa$points, type = "n", xlab = "PC1", ylab = "PC2", 
     main = "PCoA with imperviousness classes")
points(birds_pcoa$points, col = "black", bg = "orange", pch = 21)
text(birds_pcoa$points, rownames(birds_pcoa$points), cex = 0.7)
ordiellipse(birds_pcoa, envir$imperv_classes, col = 1:3, 
            draw = "polygon")
ordispider(birds_pcoa, envir$imperv_classes, col = 1:3, 
            label = TRUE)
```


# <span class='exercise'>Exercise</span>

See script on alpha diversity exercise Q 4


<br><hr><br>

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
#git2r::repository() ## uncomment if you are using GitHub
sessionInfo()
```

</details>
